---
title: "An overview of the rsi R package for retrieving satellite imagery and calculating spectral indices" 
author: "Mateusz Rydzik"
date: "2023-12-18"
slug: 
categories: 
  - rstats
tags: [rspatial, rstats]
draft: true
execute: 
  warning: false
---

# Introduction

[**rsi**](https://permian-global-research.github.io/rsi/) is a recent R package developed by [Michael Mahoney](https://www.mm218.dev/) and funded by Permian Global Research.
It offers features that simplify the process of acquiring spatial data from STAC (*SpatioTemporal Asset Catalog*) and calculating spectral indices based on such data.
A unique feature of this package is its source for the indices.
Instead of providing a static list of available formulas, **rsi** obtains them from [Awesome Spectral Indices](https://awesome-ee-spectral-indices.readthedocs.io/en/latest/): a curated repository of over 200 indices covering multiple application domains.
The combination of satellite imagery access through STAC and a constantly expanding spectral indices repository significantly simplifies remote sensing processes and creates new opportunities, including the automation of calculating spectral indices over a wide span of time and area.

# Set up

**rsi** is available in development version on [GitHub](https://github.com/Permian-Global-Research/rsi), which can be installed with [pak](https://pak.r-lib.org/).

``` {r downloading rsi}
pak::pak("Permian-Global-Research/rsi")
```

To acquire the newest Awesome Spectral Indices dataset, we can call [`spectral_indices()`](https://permian-global-research.github.io/rsi/reference/spectral_indices.html). 
Calling it without passing any arguments will result in using cached version of the tibble, which is updated automatically once a day.
This tibble contains column with indices formulas, which are using standardized band names.

``` {r loading spectral indices}
library(rsi)
asi = spectral_indices(download_indices = TRUE)
asi
``` 

# Acquiring STAC data

**rsi** provides [`get_stac_data()`](https://permian-global-research.github.io/rsi/reference/get_stac_data.html#ref-usage) function, built around the **rstac** package.
It allows to connect to specified STAC source, and access data, based on selected attributes, like geospatial area of interest, date of acquisition of imagery, collection name, etc.
For users convenience, **rsi** comes with additional functions that facilitate `get_stac_data()` function, providing necessary parameters, and limiting users input to specifying area of interest and a time frame of imagery. 
Those functions allow to access Landsat, Sentinel 1-2 imagery and digital elevation models (DEMs), available on Microsoft's Planetary Computer.

Let's start with creating our area of interest. 
`get_stac_data()` accepts `sf` objects from which it extracts boundaries, so we can load in already existing data source, or create our own.
For this example, we will create area of interest of 5000 meters around a point in San Antonio.

``` {r creating san antonio area of interest}
library(sf)
san_antonio = st_point(c(-98.491142, 29.424349))
san_antonio = st_sfc(san_antonio, crs = "EPSG:4326")
san_antonio = st_buffer(st_transform(san_antonio, "EPSG:3081"), 5000)
```

Having specified the area of interest, we can create our query.
We will start with downloading Landsat data for September and October 2023 and save it to a temporary file.

``` {r downloading landsat for sa}
sa_landsat = get_landsat_imagery(
    san_antonio,
    start_date = "2023-09-01",
    end_date = "2023-10-31",
    output_filename = tempfile(fileext = ".tif")
)
sa_landsat
```

`get_stac_data()` and its child functions return the path to downloaded image. 
As a default, they create the composite image out of median values of all avaiable images for specified time span. 
We can specify other composite functions, like `mean`, `sum`, `min` or `max`. 
We can also not use composite function at all (`composite_function = NULL`), and obtain all available images for specified time span. 
Another default argument, passed in for `mask_band`, automatically masks clouds with NAs.

``` {r downloading sentinel2 for sa}
sa_sentinel2 = get_sentinel2_imagery(
    san_antonio,
    start_date = "2023-09-01",
    end_date = "2023-10-31",
    output_filename = tempfile(fileext = ".tif")
)
```

In this example, we download Sentinel 2 images for September and October 2023, mask out clouds, and create a composite image.

Let's also obtain the DEM of our area.

``` {r downloading dem for sa}
sa_dem = get_dem(
    san_antonio,
    output_filename = tempfile(fileext = ".tif")
)
```

After downloading the images, we can load them into `SpatRaster` objects using [`rast()`](https://rdrr.io/cran/terra/man/rast.html).
**rsi** automatically assigns proper band names to it, which follow Awesome Spectral Indices standard.

``` {r loading imagery as SpatRaster}
library(terra)
sa_landsat_rast = terra::rast(sa_landsat)
sa_sentinel2_rast = terra::rast(sa_sentinel2)
sa_dem_rast = terra::rast(sa_dem)
sa_landsat_rast
```

We can plot the rasters together to compare Landsat and Sentinel images and visualize DEM.

``` {r plotting rasters}
par(mfrow = c(1, 3))
terra::plotRGB(sa_landsat_rast, r = 4, g = 3, b = 2, stretch = "lin")
terra::plotRGB(sa_sentinel2_rast, r = 4, g = 3, b = 2, stretch = "lin")
terra::plot(sa_dem_rast)
```

# Calculating spectral indices

Having downloaded images, we can use them to calculate spectral indices. 
In this example, we will observe the change in vegetation between two dates for the Swarzędz area in Poland.
Let's start by creating a new area of interest.

``` {r creating swarzedz area of interest}
swarzedz = st_point(c(17.108174, 52.405725))
swarzedz = st_set_crs(st_sfc(swarzedz), "EPSG:4326")
swarzedz = st_buffer(st_transform(swarzedz, "EPSG:2180"), 5000)
```

For the most accurate results, we will use Sentinel 2 data. 
The date span is set from July to September 2023. 
We will also pass in `composite_function = NULL`, so that all available images will be downloaded separately.

``` {r downloading sentinel2 for swarzedz}
swarzedz_sentinel2_sep = get_sentinel2_imagery(
    swarzedz,
    start_date = "2023-07-01",
    end_date = "2023-09-30",
    output_filename = tempfile(fileext = ".tif"),
    composite_function = NULL
)

swarzedz_sentinel2_sep
```

**rsi** allows for [creating custom query functions using CQL2](https://permian-global-research.github.io/rsi/articles/Downloading-data-from-STAC-APIs-using-rsi.html#using-cql2-to-refine-queries-to-stac-apis), which we can use to filter out products that do not meet our criteria. A common use case is filtering items based on their cloud coverage. For instance, if we want to download Sentinel-2 imagery with cloud cover below 25%, we can define our query function as follows:

``` {r creating query function for sentinel2 25% cloud cover}
sentinel2_25cc_qf <- function(bbox, stac_source, start_date, end_date, limit, ...) {
    geom <- rstac::cql2_bbox_as_geojson(bbox)
    datetime <- rstac::cql2_interval(start_date, end_date)
    request <- rstac::ext_filter(
        rstac::stac(stac_source),
        collection == "sentinel-2-l2a" &&
        t_intersects(datetime, {{datetime}}) &&
        s_intersects(geom, {{geom}}) &&
        platform == "Sentinel-2B" &&
        `eo:cloud_cover` < 25
        )
    rstac::items_fetch(rstac::post_request(request))
}
```

We could pass it in using the code `get_stac_data(query_function = sentinel2_25cc_qf)`. This is an optional argument, as **rsi** provides a default query function. We can then continue our work with the results stored in `swarzedz_sentinel2_sep`, selecting two images that have low cloud coverage.

``` {r selecting images for two specific dates}
swarzedz_sentinel2_07_09 = terra::rast(swarzedz_sentinel2_sep[16])
swarzedz_sentinel2_09_27 = terra::rast(swarzedz_sentinel2_sep[1])
```

**rsi**'s [calculate_indices()](https://permian-global-research.github.io/rsi/reference/calculate_indices.html) requires a data frame of indices to calculate.
Since we want to calculate [NDVI](https://en.wikipedia.org/wiki/Normalized_difference_vegetation_index), all we need to do is extract the row with NDVI's formula. 
We can use the `short_name` column for that.

``` {r calculating NDVI}
ndvi = asi[asi$short_name == "NDVI", ]
swarzedz_sentinel2_07_09_ndvi = calculate_indices(
    swarzedz_sentinel2_07_09,
    ndvi,
    output_filename = tempfile(fileext = ".tif")
)
swarzedz_sentinel2_09_27_ndvi = calculate_indices(
    swarzedz_sentinel2_09_27,
    ndvi,
    output_filename = tempfile(fileext = ".tif")
)
swarzedz_sentinel2_07_09_ndvi
```

We now can plot both rasters to see the values that were calculated for both dates.

``` {r plotting NDVI rasters}
par(mfrow = c(1, 2))
swarzedz_sentinel2_07_09_ndvi_rast = terra::rast(swarzedz_sentinel2_07_09_ndvi)
swarzedz_sentinel2_09_27_ndvi_rast = terra::rast(swarzedz_sentinel2_09_27_ndvi)
terra::plot(swarzedz_sentinel2_07_09_ndvi_rast, range = c(-1, 1))
terra::plot(swarzedz_sentinel2_09_27_ndvi_rast, range = c(-1, 1))
```

To better visualize the change in NDVI over time, we can create a difference raster.
Values above 0 indicate that NDVI was higher on October 27 than on July 9. 
Values below 0 indicate that the NDVI values decreased on October 27, compared to July 9. 
Values equal to 0 mean that NDVI values haven't changed over time.

``` {r calculating difference raster}
par(mfrow = c(1, 2))
dif = swarzedz_sentinel2_09_27_ndvi_rast - swarzedz_sentinel2_07_09_ndvi_rast
terra::plot(dif)
hist(dif, main = "", xlab = "NDVI")
```

We can save both rasters as one, using [`stack_rasters()`](https://permian-global-research.github.io/rsi/reference/stack_rasters.html) function.

``` {r stacking result rasters}
stack = stack_rasters(
  c(
    swarzedz_sentinel2_07_09_ndvi,
    swarzedz_sentinel2_09_27_ndvi
  ),
  tempfile(fileext = ".vrt")
)
stack_rast = terra::rast(stack)
names(stack_rast) = c("NDVI 07.09", "NDVI 09.27")
stack_rast
```

``` {r plotting stack rasters}
terra::plot(stack_rast, range = c(-1, 1))
```

# Conclusion

The goal of this article was to demonstrate the capabilities of **rsi** package on the most common examples in remote sensing tasks.
It's current state already shows huge potential in many remote sensing applications. 
**rsi** provides functions which simplify processes of downloading satellite imagery and calculating spectral indices.
If you want to learn more about **rsi** package, you can visit [GitHub repository of the project](https://github.com/Permian-Global-Research/rsi).

While working on this article, I found many posibble features that could be added into **rsi**.
An useful feature would be adding methods of simpler product filtering, based on cloud coverage and other parameters. 
Other feature could be focused on time series analysis, by calculating indices for each available item in selected time span.

I'm looking forward into the future developements of **rsi**. 
