---
title: "Conversions between different spatial classes in R"
author: Jakub Nowosad
date: "2021-06-17"
slug: spatial-classes-conversion
categories: [vignette]
tags: [geocompr, sf, sp, stars, raster, terra, sabre, tmap, rstats]
draft: false
bibliography: refs.bib
---

```{r, echo=FALSE, eval=FALSE}
knitr::write_bib(c("sp", "sf", "stars", "raster", "terra", "sabre", "tmap", 
                   "spDataLarge", "spData"), 
                 "content/post/2021/refs.bib")
```

The R programming language has, over the past two decades, evolved substantial spatial data analysis capabilities, and is now one of the most powerful environments for undertaking geographic research using a reproducible command line interface.
Currently, dedicated R packages allow to read spatial data and apply a plethora of different kinds of spatial methods in a reproducible fashion.

There are two main^[There are also other spatial data models, including meshes or point clouds.] spatial data models - spatial vector data and spatial raster data.
Natively R does not support spatial data and does not have a definition of spatial classes.
Therefore, there had been [a need to create R tools able to represent spatial vector and raster data ](https://geocompr.robinlovelace.net/intro.html#the-history-of-r-spatial).
Spatial classes are slightly different from regular R objects, such as data frames or matrices, as they need to not only store values, but also information about spatial locations and their [coordinate reference systems](https://geocompr.robinlovelace.net/spatial-class.html#crs-intro).

Nowadays, the most prominent packages to represent spatial vector data are **sf** [@R-sf] and its predecessor **sp** [@R-sp], however, the **terra** [@R-terra] package also has its own spatial class for vector data.
Spatial raster data can be stored as objects from **terra** [@R-terra] and its predecessor **raster** [@R-raster], or alternatively the **stars** package [@R-stars].

As you could see in [our Why R? webinar talk](/post/2020/whyr_webinar004/), the spatial capabilities of R constantly expand, but also evolve.
New packages are being developed, while old ones are modified or superseded.
In this process, new methods are created, higher performance code is added, and possible workflows are expanded.
Alternative approaches allow for a (hopefully) healthy competition, resulting in better packages.
Of course, having more than one package (with its own spatial class/es) for a vector or raster data model could be problematic, especially for new or inexperienced users.

First, it takes time to understand how different spatial classes are organized.
To illustrate this, let's read the same spatial data, `srtm.tif` from the **spDataLarge** package [@R-spDataLarge], using **raster** and **stars**.
The **raster** object:

```{r, message=FALSE}
raster_file_path = system.file("raster/srtm.tif", package = "spDataLarge")
library(raster)
srtm_raster = raster(raster_file_path)
srtm_raster
```

The **stars** object:

```{r, message=FALSE}
library(stars)
srtm_stars = read_stars(raster_file_path)
srtm_stars
```

Secondly, other packages with methods we want to use, could only accept one specific spatial class, but not the other.
For example, the current version of the **sabre** package [@R-sabre] (0.3.2) accepts objects from the **raster** package, but not ones from **terra** or **stars**^[I plan to improve that in a future release.].
The `partitions1` and `partitions2` objects are of the `RasterLayer` class from **raster**, so the `vmeasure_calc()` function works correctly.

```{r, echo=FALSE, eval=FALSE}
remotes::install_github("nowosad/sabre@e3c0f18")
```

```{r, message=FALSE}
library(sabre)
library(raster)
data("partitions1")
data("partitions2")
vmeasure_calc(partitions1, partitions2)
```

However, when the input object (representing the same spatial data!) is of the `SpatRaster` class from **terra**, the calculation results in error.

```
vmeasure_calc(partitions1_terra, partitions1_terra)
```

```
## Error in UseMethod("vmeasure_calc") : 
##  no applicable method for 'vmeasure_calc' applied to an object of class "SpatRaster"
```

Some packages, such as **tmap** [@R-tmap], accept many R spatial classes, however, this takes a lot of effort from package creators to make it possible and to maintain it.
Gladly, a number of functions exist that allow to convert between different R spatial classes.
Using them, we can work with our favorite spatial data representation, switch to some other representation just for a certain calculation, and then convert the result back into our class.
The next two sections showcase how to move between different spatial vector and raster data classes in R.

## Spatial vector data

The `world.gpkg` file from the **spData** [@R-spData] contains spatial vector data with world countries.

```{r}
world_path = system.file("shapes/world.gpkg", package = "spData")
```

Now, we can read this file, for example, as an `sf` object, and convert it into other spatial vector data classes.

```{r, message=FALSE}
library(sf)
library(sp)
library(terra)

# read as sf
world = read_sf(world_path)

# sf to sp
world_sp1 = as(world, "Spatial")

# sf to terra vect
world_terra1 = vect(world)

# sp to terra vect
world_terra2 = vect(world_sp1)

# sp to sf
world_sf2 = st_as_sf(world_sp1)

# terra vect to sf
world_sf3 = st_as_sf(world_terra1)

# terra vect to sp
world_sp2 = as(world_terra1, "Spatial")
```

In summary, `st_as_sf()` converts other classes into `sf`, `vect()` transform other classes into **terra**'s `SpatVector`, and with `as(x, "Spatial")` it is possible to get **sp**'s vectors.

```{r, echo=FALSE}
library(tibble)
library(kableExtra)
v = tribble(
  ~"FROM/TO", ~sf, ~sp, ~terra,
  "sf", "",  'as(x, "Spatial")', "vect()",
  "sp", "st_as_sf()", "", 'vect()',
  "terra", "st_as_sf()",  'as(x, "Spatial")', ""
)
kbl(v) %>% 
  column_spec(1:3, border_left = FALSE, border_right = TRUE) %>%
  column_spec(1, bold = TRUE) %>% 
  column_spec(2:4, italic = TRUE) %>% 
  row_spec(1:3, hline_after = TRUE) %>% 
  kable_styling()
```

## Spatial raster data

The `srtm.tif` file from the **spDataLarge** [@R-spDataLarge] contains a raster elevation model for the Zion National Park in the USA.

```{r}
srtm_path = system.file("raster/srtm.tif", package = "spDataLarge")
```

Now, we can read this file, for example, as a **raster** object, and convert it into other spatial vector data classes.

```{r}
library(raster)
library(stars)
library(terra)

srtm_raster1 = raster(srtm_path)

# raster to terra
srtm_terra1 = rast(srtm_raster1)

# terra to raster
srtm_raster2 = raster(srtm_terra1)

# raster to stars
srtm_stars1 = st_as_stars(srtm_raster1)

# stars to raster
srtm_raster2 = as(srtm_stars1, "Raster")

# terra to stars
srtm_stars2 = st_as_stars(srtm_terra1)

# stars to terra
srtm_terra1a = as(srtm_stars1, "SpatRaster")
```

As you can see - in most cases, we can just use one function to move from one class to another.

```{r, echo=FALSE}
library(tibble)
library(kableExtra)
v = tribble(
  ~"FROM/TO", ~raster, ~terra, ~stars,
  "raster", "",  'rast()', "st_as_stars()",
  "terra", "raster()", "", 'st_as_stars()',
  "stars", "raster()",  'as(x, "SpatRaster")', ""
)
kbl(v) %>% 
  column_spec(1:3, border_left = FALSE, border_right = TRUE) %>%
  column_spec(1, bold = TRUE) %>% 
  column_spec(2:4, italic = TRUE) %>% 
  row_spec(1:3, hline_after = TRUE) %>% 
  kable_styling()
```

## Summary

This blog post summarizes how to move between different R spatial vector and raster classes.
All of the functions mentioned above have one role: to change classes of input objects. 
They do not, however, change geometries or underlining values in the data.

Additionally, switching from the spatial vector data model to the spatial raster data model (and vice versa) is also possible. 
These operations are known as rasterization and vectorization, and they could impact spatial and nonspatial information in the input data.
To learn more about them, read [the Raster-Vector Interactions section](https://geocompr.robinlovelace.net/geometric-operations.html#raster-vector) in [Geocomputation with R](https://geocompr.robinlovelace.net/).

## References
