---
title:  "Specialized R packages for spatial machine learning: An introduction to RandomForestsGLS, spatialRF, and meteo"
author:
  - name: Jakub Nowosad
    url: https://jakubnowosad.com
    orcid:  0000-0002-1057-3721
date: "2025-06-25"
slug: sml-bp5
categories:
  - rstats
  - sml
tags: [RandomForestsGLS, spatialRF, meteo, machine-learning, spatial-machine-learning]
draft: true
mermaid-format: png
knitr:
  opts_chunk:
    message: false
---

::: {.callout-note appearance="simple"}

This is the fifth part of a blog post series on spatial machine learning with R.

You can find the list of other blog posts in this series [in part one](/post/2025/sml-bp1/).

:::


This document provides an overview of three R packages, **RandomForestsGLS**, **spatialRF**, and **meteo**, that implement spatial machine learning methods, but are outside of standard machine learning frameworks like **caret**, **tidymodels**, or **mlr3**.^[There are several other packages that implement spatial machine learning methods, such as **SpatialML** or **ENMeval**, but they are not discussed here. Moreover, new methods and packages in this domain are being developed all the time, so this list is definitely not exhaustive.]

All of the examples below use the same dataset, which includes the temperature measurements in Spain, a set of covariates, and the spatial coordinates of the temperature measurements.

```{r misc-sml-1 }
#| message: false
#| warning: false
spain <- sf::read_sf(
    "https://github.com/LOEK-RS/FOSSGIS2025-examples/raw/refs/heads/main/data/spain.gpkg"
)
covariates <- terra::rast(
    "https://github.com/LOEK-RS/FOSSGIS2025-examples/raw/refs/heads/main/data/predictors.tif"
)
temperature <- sf::read_sf(
    "https://github.com/LOEK-RS/FOSSGIS2025-examples/raw/refs/heads/main/data/temp_train.gpkg"
)

temperature <- terra::extract(covariates, temperature, bind = TRUE) |>
    sf::st_as_sf()
```

# RandomForestsGLS

The **RandomForestsGLS** (<https://doi.org/10.21105/joss.03780>) package implements [the Generalised Least Square (GLS) based Random Forest (RF-GLS) algorithm](https://doi.org/10.1080/01621459.2021.1950003).^[Quoting the authors "RF-GLS extends RF in the same way generalized least squares (GLS) fundamentally extends ordinary
least squares (OLS) to accommodate for dependence in linear models."]
This approach is designed for spatial data modeling as it accounts for spatial dependencies in the data by:

1. Using a global dependency-adjusted split criterion and node representatives instead of the classification and regression tree (CART) criterion used in standard RF models
2. Applying contrast resampling rather than the bootstrap method used in a standard RF model
3. Employing residual kriging with covariance modeled using a Gaussian process framework

The package provides four functions:

1. `RFGLS_estimate_spatial()` for estimation in spatial data
2. `RFGLS_predict()` for prediction of the mean function
3. `RFGLS_predict_spatial()` for prediction of the spatial response
4. `RFGLS_estimate_timeseries()` for estimation in time series data (not discussed here)

The package has rather unintuitive syntax and requires the data to be in a specific format. 
We need to provide the coordinates of the data (a matrix), the response variable (a vector), and the covariates (a matrix).
In the example below, I limited the covariate matrix to the variables that are not spatial proxies.

```{r misc-sml-2 }
library(RandomForestsGLS)

coords <- sf::st_coordinates(temperature)
temp_response <- temperature$temp

temperature_df <- sf::st_drop_geometry(temperature)
covariate_names <- colnames(temperature_df)[2:(ncol(temperature_df) - 7)]
covariate_matrix <- as.matrix(temperature_df[, covariate_names])
```

I also split the data into training and testing sets based on created random indices.

```{r misc-sml-3 }
set.seed(2025 - 01 - 30)
train_idx <- sample(1:nrow(coords), floor(nrow(coords) * 0.7))
```

The `RFGLS_estimate_spatial()` function is used to fit the RF-GLS model.
Here, we customize the number of trees to 100, but the function has many other parameters that can be adjusted.

```{r misc-sml-4 }
estimation_result <- RFGLS_estimate_spatial(
    coords = coords[train_idx, ],
    y = temp_response[train_idx],
    X = covariate_matrix[train_idx, ],
    ntree = 100
)
str(estimation_result)
```

The result is a list with seven elements: a matrix of zero-indexed resamples, a matrix of predictions (ntest x ntree), a vector of predicted values, the covariate matrix, the response variable, the coordinates matrix, and the RF-GLS object.

Now, we can use the fitted model to predict the mean function (`RFGLS_predict()`) or the spatial response (`RFGLS_predict_spatial()`).
The difference (as far as I understand) is that the former returns the mean prediction, while the latter uses the spatial coordinates in addition to the covariates to predict the spatial response.

The first function returns a list with two elements: a matrix of predictions (ntest x ntree) and a vector of predicted values, while the second function returns a list with just one element: a vector of predicted values.
Just note that the predictions by the `RFGLS_predict()` are named `"predicted"` and the predictions by the `RFGLS_predict_spatial()` are named `"prediction"`.

```{r misc-sml-5 }
#| warnings: false
prediction_result <- RFGLS_predict(
    RFGLS_out = estimation_result,
    Xtest = covariate_matrix[-train_idx, ]
)
prediction_result_spatial <- RFGLS_predict_spatial(
    RFGLS_out = estimation_result,
    coords.0 = coords[-train_idx, ],
    Xtest = covariate_matrix[-train_idx, ]
)
plot(prediction_result$predicted, prediction_result_spatial$prediction)
```

The final results of these two approaches are v. similar, but not identical.

Now, let's predict the models' results on the whole dataset.

```{r misc-sml-6 }
covariate_coords_r <- terra::crds(covariates)
covariate_matrix_r <- as.matrix(covariates)
covariate_matrix_r <- covariate_matrix_r[, covariate_names]

pred_s <- RFGLS_predict_spatial(
    RFGLS_out = estimation_result,
    coords.0 = covariate_coords_r,
    Xtest = covariate_matrix_r
)

pred_r <- terra::setValues(covariates[[1]], pred_s$prediction)
names(pred_r) <- "prediction"
terra::plot(pred_r)
```

# spatialRF

The **spatialRF** (<https://blasbenito.github.io/spatialRF/>) package's aim is to provide a minimal code interface to fit spatial regression models with Random Forest.
The internal calculations are based on three general methods to generate spatial predictors from the distance matrix of the data points: Distance matrix columns as explanatory variables (Hengl et al. 2018), Moran’s Eigenvector Maps (Dray, Legendre, and Peres-Neto 2006) and PCAs.
The **ranger** package is used here internally to fit the Random Forest model.

This package also requires the data to be in a specific format.
We need to provide the data as a data frame with the dependent variable, including spatial coordinates, and the distance matrix: a matrix with the distances among the records in the data frame.

```{r misc-sml-7 }
library(spatialRF)
library(sf)

coordinates <- st_coordinates(temperature)
colnames(coordinates) <- c("x", "y")
coordinates <- as.data.frame(coordinates)

temperature_df <- st_drop_geometry(temperature)
temperature_df$x <- coordinates[, 1]
temperature_df$y <- coordinates[, 2]

distance_matrix <- as.matrix(dist(temperature_df[2:(ncol(temperature_df) - 9)]))
```

We also need to define the dependent variable and the predictor variables.

```{r misc-sml-8 }
response_name <- "temp"
covariate_names <- colnames(temperature_df)[2:(ncol(temperature_df) - 9)]
```

Finally, we can fit the models using one of the methods provided by the package.
The package has 10 methods implemented, nine of which are based on the three components:^[See `?rf_spatial` for more details. Also, the 10th method is `"hengl"` directly following the approach by Hengl et al. (2018).]

1. The method to generate spatial predictors (`"hengl"`, `"mem"`, or `"pca"`)
2. The method to rank spatial predictors (`"moran"` or `"effect"`)
3. The method to select spatial predictors (`"sequential"` or `"recursive"`)

The main function of this package is `rf_spatial()`, which fits the Random Forest model with spatial predictors.
Here, an example using Moran's Eigenvector Maps method to generate spatial predictors, Moran's I to rank them, and sequential selection of the predictors is shown.

```{r misc-sml-9 }
rf_spatial_moran <- rf_spatial(
    data = temperature_df,
    dependent.variable.name = response_name,
    predictor.variable.names = covariate_names,
    distance.matrix = distance_matrix,
    distance.thresholds = 0,
    method = "mem.moran.sequential",
    n.cores = 1
)
rf_spatial_moran
```

The `rf_spatial()` returns a **ranger** model with several new slows, most importantly `residuals` that contain information about the residuals, and `spatial` that contains information about the selected spatial predictors and the method used to select them.
Printing the model object provides a summary of the model, including its parameters, model performance, information on model residuals, and variable importance. 

The **spatialRF** package also provides a set of additional functions.
It includes a function for reducing multicollinearity in the predictors and removing redundant spatial predictors (`filter_spatial_predictors()`); or finding promising variable interactions (`the_feature_engineer()`):

```{r misc-sml-10 }
interactions <- the_feature_engineer(
    data = temperature_df,
    dependent.variable.name = response_name,
    predictor.variable.names = covariate_names,
    xy = coordinates,
    importance.threshold = 0.50, # uses 50% best predictors
    cor.threshold = 0.60, # max corr between interactions and predictors
    seed = 2025 - 01 - 30,
    repetitions = 100,
    verbose = TRUE
)
```

The `rf_evaluate()` function allows the evaluation of the model using spatial cross-validation.

```{r misc-sml-11 }
rf_eval <- rf_evaluate(
    model = rf_spatial_moran,
    xy = coordinates,
    repetitions = 30,
    training.fraction = 0.75,
    metrics = "rmse",
    seed = 2025 - 01 - 30,
    verbose = TRUE
)
rf_eval
```

The `rf_importance()` function allows for visualizing the variable importance of the model.
```{r misc-sml-12 }
rf_imp <- rf_importance(
    rf_spatial_moran,
    xy = coordinates
)
rf_imp
```

The `mem()` function generates Moran Eigenvector Maps (MEM) from a distance matrix.^[`mem_multithreshold()` function allows for generating MEMs for multiple distance thresholds.]

```{r misc-sml-13 }
mem1 <- mem(distance.matrix = distance_matrix)
```

The package also contains a set of custom plot functions. 
One example is the `plot_response_curves()` function, which allows for the visualization of the response curves of the model.

```{r misc-sml-14 }
plot_response_curves(rf_spatial_moran)
```

Additional interesting functions allow for tuning the model parameters (`rf_tuning()`) or comparing several models (`rf_compare()`).
A complete list of this package's functions is available at <https://blasbenito.github.io/spatialRF/reference/index.html>.

The final prediction can be made using the `predict()` function from the **terra** package.

```{r misc-sml-15 }
pred_srf <- terra::predict(covariates, rf_spatial_moran)
terra::plot(pred_srf[[1]])
```

# meteo

Random Forest Spatial Interpolation (RFSI, Sekulić et al. (2020) <doi:10.3390/rs12101687>) is implemented in the **meteo** package.
RFSI enhances traditional Random Forest by explicitly incorporating spatial autocorrelation through additional covariates. 
These covariates include (1) observations from the nn nearest locations and (2) their respective distances to the target location. 
Predictions follow the same principle, using the nearest observed values and distances to improve spatial accuracy.

```{r misc-sml-16}
library(meteo)
```

This package allows to directly work with spatial R objects.
First, we need to define our formula.

```{r misc-sml-17}
response_name <- "temp"
covariate_names <- colnames(temperature)[2:(ncol(temperature) - 8)]
fo <- as.formula(paste(
    response_name,
    "~",
    paste(covariate_names, collapse = " + ")
))
```

Next, we use the main function in this package, `rfsi()`, to fit the RFSI model.
The function requires the formula, the data, and the number of nearest observations used in the model.
Additional arguments can be set, including the number of CPUs to use, the progress bar, and arguments passed to the **ranger** function, such as `seed`, `num.trees`, and `mtry`.

```{r misc-sml-18}
rfsi_model <- rfsi(
    formula = fo,
    data = temperature,
    n.obs = 5, # number of nearest observations
    cpus = parallel::detectCores() - 1,
    progress = FALSE,
    importance = "impurity",
    seed = 42,
    num.trees = 250,
    mtry = 5
)
rfsi_model
```

The outcome of this function is a standard `ranger` object.

Another function, `cv.rfsi()`, allows for spatial (leave-location-out) cross-validation of the RFSI model.
It requires the formula, the data, the tuning grid, the type of cross-validation, the number of folds, and few other arguments.
The `tgrid` argument is a data frame with the tuning parameters to be tested that may include `mtry`, `num.trees`, `n.obs`, and `sample.fraction`.
Here, we will only tune the `mtry` parameter.

```{r misc-sml-19}
#| message: false
#| warning: false
rfsi_model_cv <- cv.rfsi(
    formula = fo,
    data = temperature,
    tgrid = expand.grid(mtry = 3:22),
    tune.type = "LLO", # Leave-Location-Out CV
    k = 5, # number of folds
    seed = 42,
    acc.metric = "RMSE", # R2, CCC, MAE
    output.format = "sf", # "data.frame", # "SpatVector",
    cpus = parallel::detectCores() - 1,
    progress = FALSE,
    importance = "impurity"
) # ranger parameter
rfsi_model_cv
acc.metric.fun(rfsi_model_cv$obs, rfsi_model_cv$pred, "RMSE")
```

The result is an object of the desired output format (here, a `sf` object) with the observed and predicted values.
The `acc.metric.fun()` function can be then use to calculate the accuracy metric (here, RMSE) between the observed and predicted values.

Finally, we can use the `pred.rfsi()` function to predict the model on new data, with a selected output format (here, `SpatRaster`).

```{r misc-sml-20}
# Prediction on new data
rfsi_prediction <- pred.rfsi(
    model = rfsi_model,
    data = temperature,
    obs.col = "temp",
    newdata = covariates,
    output.format = "SpatRaster", # "sf", # "SpatVector"
    cpus = parallel::detectCores() - 1,
    progress = FALSE
)
terra::plot(rfsi_prediction)
```


::: {.content-hidden}

## ENMeval

<https://jamiemkass.github.io/ENMeval/articles/ENMeval-2.0-vignette.html>

Out-of-scope: a focus on the occurrence records (ecological niche models/species distribution models)

## sits

<https://github.com/e-sensing/sits>

Out-of-scope: a focus on spatiotemporal data cubes

## SpatialML {.}

SpatialML implements a spatial extension of the random forest algorithm (Georganos et al. (2019) <doi:10.1080/10106049.2019.1595177>) based on the **randomForest**, **ranger**, and **caret** packages.
This extension is called Geographical Random Forest (GRF).
It extends traditional Random Forest by incorporating spatial heterogeneity through local sub-models, where each data point is used to train a localized RF using nearby observations, controlled by an adaptive kernel. 

This package is not described in details here (nor the code is run) as it is aimed at predictions of rather small datasets (e.g., administrative polygons) -- the prediction of the example data takes a long time (~45 minutes).

```{r misc-sml-21}
#| eval: false
library(SpatialML)
```

```{r misc-sml-22}
#| eval: false
coordinates <- sf::st_coordinates(temperature)
temperature_df <- sf::st_drop_geometry(temperature)

# define formula
response_name <- "temp"
covariate_names <- colnames(temperature_df)[2:(ncol(temperature_df) - 7)]
fo <- as.formula(paste(
    response_name,
    "~",
    paste(covariate_names, collapse = " + ")
))

sml <- grf(
    fo,
    dframe = temperature_df,
    bw = 8,
    kernel = "adaptive",
    coords = coordinates
)
```

```{r misc-sml-23}
#| eval: false
# 1
covariates_df <- terra::as.data.frame(covariates)
covariates_df <- na.omit(covariates_df)
sml_prediction <- predict.grf(
    sml,
    covariates_df,
    x.var.name = "X",
    y.var.name = "Y"
)

# 2
system.time({
    sml_prediction <- terra::predict(
        covariates,
        sml,
        fun = predict.grf,
        x.var.name = "X",
        y.var.name = "Y",
        na.rm = TRUE
    )
})
```

:::

::: {.callout-note appearance="simple"}

This blog post was originally written as a supplement to the poster "An Inventory of Spatial Machine Learning Packages in R" presented at the FOSSGIS 2025 conference in Muenster, Germany.
The poster is available at <https://doi.org/10.5281/zenodo.15088973>.

:::